---
title: "Block penalty simulations analysis"
author: Ahmed Hasan
output: pdf_notebook
---

```{r setup, include=FALSE, echo=FALSE}
require('knitr')
opts_knit$set(root.dir = '../../')
```


```{r}
library(tidyverse)
library(fs)
library(magrittr, warn.conflicts = FALSE)
library(glue)
```

# Loading in data

```{r}
load_ldhelmet_files <- function(window, flank, block) {
  # window and flank in kb
  
  full_dir <- glue('data/macs-runs/ldhelmet_{window}_{flank}/block_{block}')
  
  out <- dir_ls(full_dir) %>% 
    map_dfr(., read_csv, col_types = cols(), .id = 'name') %>% 
    select(-chr) %>%
    mutate(name = str_extract(name, '[0-9]\\.[0-9]+_10[0-9]')) %>% # get fname
    separate(name, into = c('rho', 'run'), sep = '_', convert = TRUE) %>%  # pull rho + run out of fname
    rename(start = block_start, end = block_end)
  return(out)
}

block_5 <- load_ldhelmet_files(2000, 40000, 5)
block_10 <- load_ldhelmet_files(2000, 40000, 10)
block_50 <- load_ldhelmet_files(2000, 40000, 50)
block_100 <- load_ldhelmet_files(2000, 40000, 100)

block_5
```

# Exploring the data

## Overall distributions

What does the distribution of rho look like in block = 10?

```{r}
ggplot(block_10, aes(x = block_rate)) +
  geom_histogram() +
  theme_classic()
```

Hotspots were supposed to be placed as follows:

```
0.0500	0.0520	10.00000
0.1786	0.1806	10.00000
0.3071	0.3091	20.00000
0.4357	0.4377	20.00000
0.5643	0.5663	40.00000
0.6929	0.6949	40.00000
0.8214	0.8234	60.00000
0.9500	0.9520	60.00000
```

Where the first two columns represent chunks of the sequence (over [0, 1])
and the third one represents fold increase over baseline rho (i.e. value in filename)

These can be stuck into a data frame for easy comparison:

```{r}
hotspots <- data.frame(
  prop_start = c(0.0500, 0.1786, 0.3071, 0.4357, 0.5643, 0.6929, 0.8214, 0.9500),
  prop_end = c(0.0520, 0.1806, 0.3091, 0.4377, 0.5663, 0.6949, 0.8234, 0.9520),
  rate_increase = c(10, 10, 20, 20, 40, 40, 60, 60)
) %>%
  mutate(exact_start = prop_start * 1e6,
         exact_end = prop_end * 1e6) %>% 
  select(-starts_with('prop')) %>% 
  mutate(start = round(exact_start / 1000) * 1000,
         end = round(exact_end / 1000) * 1000) %>% 
  select(start, end, everything())
```

## Background rho estimation

Which block penalty gives the most accurate background rho estimates? (not including hotspots)

```{r}

non_overlapping <- function(df, windowsize) {
  df %<>%
    mutate(div = floor(start / windowsize), div2 = lead(div)) %>%
    filter(div == div2) %>%
    select(-contains('div'))
  return(df)
}

# starting with block 5
# quick and dirty visualization
block_5 %>%
  anti_join(hotspots, by = c('start', 'end')) %>% 
  ggplot(aes(x = block_rate)) +
  geom_histogram() +
  geom_vline(aes(xintercept = rho)) +
  facet_wrap(~ rho) +
  theme_classic()

block_5 %>% 
  anti_join(hotspots, by = c('start', 'end')) %>% 
  non_overlapping(2000) %>% 
  group_by(rho) %>% 
  summarise(estimate = mean(block_rate)) %>% 
  mutate(obs_exp = estimate / rho)
  
```

Expanding to all datasets:

```{r}
d_all <- list(block_5, block_10, block_50, block_100)
names(d_all) <- c(5, 10, 50, 100)

d_all %>% 
  map_dfr(~ anti_join(., hotspots, by = c('start', 'end')) %>% 
            non_overlapping(2000) %>% 
            group_by(rho) %>% 
            summarise(estimate = mean(block_rate)) %>% 
            mutate(obs_exp = estimate / rho),
          .id = 'block') %>% 
  mutate(block = as.numeric(block),
         rho = forcats::fct_relevel(
           as.factor(rho), 
           levels = c('1e-04', '0.001', '0.01', '0.1', '1', '2.5'))
         ) %>% 
  ggplot(aes(x = rho, y = obs_exp, color = block)) +
  geom_point() +
  geom_line() + 
  geom_hline(yintercept = 1) +
  theme_classic()
```

Looks like block = 100 is our best bet here. 


## Hotspot power estimation

power = 1 - prob of false negatives (known hotspots not detected as hotspots)

For each rho -> 10 runs, with 8 hotspots each -> max of 80 hotspots per background rho per block

20 hotspots per rate increase value per rho (group by rho, rate_increase and tally)


```{r}
d_all %>% 
  map_dfr(~ inner_join(., hotspots, by = c('start', 'end')) %>% 
  filter(rate_ratio >= rate_increase) %>% 
  group_by(rho, rate_increase) %>% 
  tally() %>% 
  mutate(power = n / 20), # how many of the 20 were correctly called as hotspots
  .id = 'block') %>% 
  ggplot(aes(x = rate_increase, y = power, color = block)) +
  geom_point() + geom_line() +
  facet_wrap(~ rho)
```

Expected:

```{r}
block_5 %>% 
  inner_join(hotspots, by = c('start', 'end')) %>% 
  # filter(rate_ratio >= rate_increase) %>% 
  group_by(rho, rate_increase) %>% 
  tally()
```

Creating a generalized function to repeat the above operation for different window/flank sizes:

```{r}
hotspot_power <- function(window, flank, verbose = FALSE) {
  # read in files
  dfs_all <- list()
  blocks <- c(5, 10, 50, 100)
  for (i in seq_along(blocks)) {
    dfs_all[[i]] <- load_ldhelmet_files(window, flank, blocks[i])
  }
  names(dfs_all) <- blocks
  
  # hotspot counts
  div <- window / 2
  hotspot_df <- data.frame(
    prop_start = c(0.0500, 0.1786, 0.3071, 0.4357, 0.5643, 0.6929, 0.8214, 0.9500),
    prop_end = c(0.0520, 0.1806, 0.3091, 0.4377, 0.5663, 0.6949, 0.8234, 0.9520),
    rate_increase = c(10, 10, 20, 20, 40, 40, 60, 60)
  ) %>%
    mutate(exact_start = prop_start * 1e6,
           exact_end = prop_end * 1e6) %>% 
    select(-starts_with('prop')) %>% 
    mutate(start = round(exact_start / div) * div,
           end = round(exact_end / div) * div) %>% 
    select(start, end, everything())
  
  dfs_all %<>%
    map_dfr(
      ~ inner_join(., hotspot_df, by = c('start', 'end')) %>% 
        filter(rate_ratio >= rate_increase) %>% 
        group_by(rho, rate_increase) %>% 
        tally() %>% 
        mutate(power = n / 20),
      .id = 'block'
    )
  if (verbose == TRUE) {
    message('done')
  }
  return(dfs_all)
}

```




































